---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

packages <- c("tidyr", "knitr", "ggplot2", "ggmap", "maps", "kableExtra", "tibble")
install.packages(setdiff(packages, rownames(installed.packages())), repos = "http://cran.us.r-project.org")

library(ggplot2)
library(ggmap)
library(maps)
library(dplyr)
library(tidyr)
library(knitr)
library(tibble)
# library(kableExtra)

theme_set(theme_light())
```

# Cyclistic

- Anar Seyf (anar.seyf@gmail.com)
- October 2021
- Capstone project | Google Data Analytics course #8 | Coursera

```{r}
agg_daily_computed_weekdays <- read.csv("agg_daily_computed_weekdays.csv")
agg_daily <- read.csv("agg_daily_duration_count.csv")
agg_hourly <- read.csv("agg_hourly_status.csv")
agg_monthly <- read.csv("agg_monthly_status.csv")
agg_weekly_status <- read.csv("agg_weekly_status.csv")

weekday_names_full <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
weekday_names_short <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
```

## Data

Basic stats, etc.


## Analysis


### 1. Members ride more often, casual users ride for longer.

Overall stats TODO. Mini charts TODO. February outliers. Summer months #C > #M.

#### Average ride duration (minutes)

```{r}

wide <- agg_monthly %>%
  select(status, month_start, duration) %>%
  spread(month_start, duration)

month_starts <- colnames(wide)[2:13]
colnames(wide) <- c("Status", months(as.Date(month_starts), abbreviate = TRUE))

wide[, -1] <- round(wide[, -1])

wide %>%
  kable(format.args = list(big.mark = ","))
# kable_styling(full_width = FALSE)
```

#### Average daily rides (count)

```{r}

wide <- agg_monthly %>%
  select(status, month_start, avg_count) %>%
  spread(month_start, avg_count)

month_starts <- colnames(wide)[2:13]
colnames(wide) <- c("Status", months(as.Date(month_starts), abbreviate = TRUE))

wide[, -1] <- round(wide[, -1])

wide %>%
  kable(format.args = list(big.mark = ","))
```


### 2. Members ride all week, casual riders prefer weekends.

Members' usage remains nearly flat, falling on Sunday. Casual users do most riding starting on Friday and through the weekend, peaking on Saturday.


### By weekday


#### Average daily duration (minutes)

```{r, fig.align='left'}
wide <- agg_weekly_status %>%
  select(-avg_count) %>%
  spread(dayofweek, avg_duration)

colnames(wide) <- c("Status", weekday_names_short)

wide %>% kable()
```

#### Average daily rides (count)

```{r}

wide <- agg_weekly_status %>%
  select(-avg_duration) %>%
  spread(dayofweek, avg_count)

colnames(wide) <- c("Status", weekday_names_short)

wide %>%
  kable(format.args = list(big.mark = ","))
```


#### Hourly

```{r, fig.width=4, fig.height=4}
agg_hourly %>%
  ggplot(aes(
    x = hour,
    y = factor(dayofweek, levels = unique(rev(dayofweek))),
    color = IS_MEMBER, size = COUNT
  )) +
  # scale_color_brewer(palette = "Pastel1") +
  geom_point(position = position_dodge(width = 0.5), alpha = 0.8) +
  theme(legend.position = "none")
```

```{r, fig.width=8, fig.height = 3}
agg_hourly %>%
  mutate(log_count = log(COUNT)) %>%
  ggplot(aes(
    x = dayofweek,
    y = factor(hour, levels = unique(rev(hour))),
    fill = log_count
  )) +
  geom_tile(color = "grey", position = position_dodge(width = 0.1)) +
  scale_fill_viridis_c(option = "plasma") +
  #  scale_fill_steps2(low = "white", mid = "yellow", high = "red", midpoint = 3) +
  facet_wrap(~IS_MEMBER)
```

## The Graphs


#### Monthly by weekday

```{r, eval=FALSE}
plot <- agg_daily_computed_weekdays %>%
  group_by(year_month, dayofweek, IS_MEMBER) %>%
  summarize(AVG_COUNT = mean(COUNT), AVG_DURATION = mean(AVG_DURATION))

plot %>%
  ggplot(aes(x = dayofweek, y = AVG_COUNT, color = IS_MEMBER, size = AVG_DURATION)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~year_month, ncol = 3)
```


#### Weekly Counts and Average Duration

Full year, weekly.
```{r, eval=FALSE, echo=FALSE, fig.width=8, fig.height=2}
head(agg_daily_computed_weekdays, n = 2)

agg_daily_computed_weekdays %>%
  ggplot(aes(
    x = factor(weeknum, levels = unique(weeknum)),
    y = COUNT,
    fill = IS_MEMBER
  )) +
  geom_bar(stat = "identity", alpha = 0.8) +
  theme(legend.position = "none")
```

```{r, eval=FALSE, fig.width=8, fig.height=2}
# y = factor(weeknum, levels = unique(rev(weeknum)))
agg_daily_computed_weekdays %>%
  # select(-STDDEV_DURATION, -MIN_DURATION, -MAX_DURATION) %>%
  group_by(weeknum, IS_MEMBER) %>%
  summarize(WEEKLY_AVG_DURATION = mean(AVG_DURATION)) %>%
  ggplot(aes(
    x = factor(weeknum, levels = unique(weeknum)),
    y = WEEKLY_AVG_DURATION,
    fill = IS_MEMBER
  )) +
  geom_bar(stat = "identity", alpha = 0.8) +
  theme(legend.position = "none")
```

Note: **not weighted average** for the graph above.

#### Monthly by bike type

```{r, eval=FALSE}
agg_monthly_type_status <- read.csv("agg_monthly_type_status.csv")
head(agg_monthly_type_status, n = 2)

agg_monthly_type_status %>%
  ggplot(aes(x = is_member, y = total_count, fill = type)) +
  geom_col() +
  facet_wrap(~year_month, ncol = 3)
```

## The Weather

#### Weekly

**TODO** — either use a weather-only dataset or show trips alongside weather.

```{r, eval=FALSE, fig.width=8, fig.height=2}
agg_weekly_weather <- read.csv("agg_weekly_weather.csv")
# head(agg_weekly_weather, n = 3)

d <- agg_weekly_weather %>%
  filter(is_member == "true") %>%
  filter(weekly_rain_total > 0) %>%
  select(-is_member)

p <- ggplot(data = d, aes(x = year_week, y = weekly_temp_avg, group = 1))

p +
  geom_ribbon(aes(
    ymin = weekly_temp_min, ymax = weekly_temp_max,
    alpha = 0.8
  )) +
  geom_line() +
  geom_point(aes(y = 0, size = weekly_rain_total, color = weekly_rain_total, alpha = 0.8))
```

**TODO: Recompute** on new daily aggregates.

```{r, eval=FALSE}
weather <- read.csv("agg-daily-trips-weather.csv")
weather %>%
  select(-wind_2min_peak, -temp_min, -temp_max) %>%
  head(n = 2)
```


#### Temperature

In degrees C.

```{r, eval=FALSE, out.width='45%', out.aspect=0.3}

ggplot(data = weather, aes(x = temp_avg, y = COUNT, color = IS_MEMBER, size = AVG_DURATION)) +
  geom_point(alpha = 0.8)
```
```{r, eval=FALSE, out.width='45%', out.aspect=0.3}
ggplot(data = weather, aes(x = temp_avg, fill = IS_MEMBER, size = AVG_DURATION)) +
  geom_histogram(binwidth = 1, alpha = 0.8)
```

```{r, eval=FALSE, out.width='45%'}
weather %>%
  ggplot(aes(x = temp_avg, y = COUNT, fill = IS_MEMBER, size = AVG_DURATION)) +
  geom_col(position = "fill", alpha = 0.8) +
  scale_x_binned() +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "none")
```


#### Rain

Rain (mm) vs ride count

```{r, eval=FALSE, out.width='45%', out.aspect=0.3}
weather %>%
  filter(rain <= 20) %>%
  ggplot(aes(x = rain, fill = IS_MEMBER, size = AVG_DURATION)) +
  # scale_fill_brewer(palette = "Pastel1") +
  geom_histogram(binwidth = 1, alpha = 0.8)
```
```{r, eval=FALSE, out.width='45%', out.aspect=0.3}
weather %>%
  filter(rain <= 20) %>%
  ggplot(aes(x = rain, y = COUNT, fill = IS_MEMBER, size = AVG_DURATION)) +
  geom_col(position = "fill", alpha = 0.8) +
  scale_x_binned() +
  scale_y_continuous(labels = scales::percent) +
  # scale_fill_brewer(palette = "Pastel1") +
  theme(legend.position = "none")
```


#### Wind

Wind (m/s) vs ride count.

Second graph is scaled to 100% to show member vs casual share.

```{r, eval=FALSE, out.width='30%'}
weather %>%
  filter(wind_avg <= 10) %>%
  ggplot(aes(x = wind_avg, y = COUNT, fill = IS_MEMBER, size = AVG_DURATION)) +
  geom_col(alpha = 0.8) +
  # scale_fill_brewer(palette = "Pastel1") +
  scale_x_binned()
```
```{r, eval=FALSE, out.width='30%'}
weather %>%
  filter(wind_avg <= 10) %>%
  ggplot(aes(x = wind_avg, y = COUNT, fill = IS_MEMBER, size = AVG_DURATION)) +
  geom_col(position = "fill", alpha = 0.8) +
  scale_x_binned() +
  # scale_fill_brewer(palette = "Pastel1") +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "none")
```

## The Geography

#### Stations

**1288** stations. Following: **casual vs members**, **8am vs 4pm**, **Monday-Friday**.

Each dot is a bike docking station. Stations highlighed **yellow-red** have more arrivals than departures for the given hour, suggesting an influx of bike traffic at that location.

```{r, eval=FALSE}
stations_diff <- read.csv("stations_diff_8am_noon_4pm.csv")

stations_diff <- stations_diff %>% arrange(diff)

index <- stations_diff$diff > 0
stations_diff$log_diff[index] <- log10(stations_diff$diff[index])
stations_diff$log_diff[stations_diff$diff <= 0] <- NA

color_scale <- scale_color_steps(low = "#ffff17", high = "#e90505", na.value = "#072d66")

# map11 <- get_map(location = "Chicago", zoom = 11, maptype = "toner")
map12 <- get_map(location = "Chicago", zoom = 12, maptype = "toner")

alpha <- 0.6
darken <- 0.4
```

#### Friday, 4pm, casual vs members

```{r, eval=FALSE, fig.width = 8, fig.height = 4}
heatmap_large <- stations_diff %>%
  filter(dayofweek == 5) %>%
  filter(hour == 16)

ggmap(map12, darken = darken) +
  # ggplot() +
  geom_point(data = heatmap_large, aes(x = lng, y = lat, color = log_diff), alpha = alpha) + color_scale +
  facet_wrap(~is_member)
```

#### 8am, Monday-Sunday, casual vs members
```{r, eval=FALSE, fig.width=10, fig.height=4}
heatmap_member <- stations_diff %>%
  filter(hour == 8)

ggmap(map12, darken = darken) +
  # ggplot() +
  geom_point(data = heatmap_member, aes(x = lng, y = lat, color = log_diff), alpha = alpha) + color_scale +
  facet_grid(is_member ~ dayofweek)
```

#### 4pm, Monday-Sunday, casual vs members
```{r, eval=FALSE, fig.width=10, fig.height=4}
heatmap_casual <- stations_diff %>%
  filter(hour == 16)

ggmap(map12, darken = darken) +
  # ggplot() +
  geom_point(data = heatmap_casual, aes(x = lng, y = lat, color = log_diff), alpha = alpha) + color_scale +
  facet_grid(is_member ~ dayofweek)
```

